---
title: "Spatial Entropy"
#date: '`r format(Sys.time(), "%Y-%m-%d")`'
output:
  html_document: 
    code_folding: hide
    toc: true
    toc_float: true
hitheme: tomorrow
highlighter: highlight.js
---


```{r setup , include=FALSE}
library(survival)
library(survminer)
library(dplyr)
library(tidyverse)
library(readxl)
library(janitor)
library(skimr)
library(phenoptr)
library(phenoptrReports)
library(rtree)
library(data.table)
library(pivottabler)
library(ggpubr)
library(zoo)
library(lubridate)
library(reReg)
library(spatstat)
library(refund)
library(mgcv)
library(scam)
library(SpatEntropy)
library(fdapace)
library(plot3D)
library(refund)
library(refund.shiny)

knitr::opts_chunk$set(echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)

theme_set(theme_bw() + theme(legend.position = "bottom"))
```


# Input data
## Load source functions
```{r}
source_dir = "~/Documents/Imaging project/Vectra /" # keep the space after Vectra
source(paste0(source_dir, "ref_functions.R"))
```

## Read in lung cancer data
```{r, echo=FALSE, warning=FALSE}
setwd("~/Documents/Imaging project/Vectra /Lung_survival_ouput")
#load("Lung_workspace.RData")
load("lung_slide_analysis.RData")
```

```{r, warning=FALSE, echo = FALSE}
mhcii <- read_csv(paste0(CDir, "HLADR_status.csv"))[,-1]
colnames(mhcii)[1] = "Sample Name"
lung_clinical_data = left_join(lung_clinical_data, mhcii, by = "Sample Name")
```

# visualize images based on survival
 
```{r, warning=FALSE, echo = FALSE, message=FALSE}
n_surv = 15
event_id = which(lung_clinical_data$`Censor (0=Alive, 1=Dead)` == 1)
censor_id = which(lung_clinical_data$`Censor (0=Alive, 1=Dead)` == 0)
#best_surv = event_id[order(lung_clinical_data$`Survival (days)`[event_id], decreasing = T)[1:n_surv]]
best_surv = censor_id[order(lung_clinical_data$`Survival (days)`[censor_id], decreasing = T)[1:n_surv]]
worst_surv = event_id[order(lung_clinical_data$`Survival (days)`[event_id])[1:n_surv]]

plt_group = head(best_surv, 12)
df <- NULL
for(i in plt_group){
   
temp = slide_idx[[i]] 
cell_rm = c(which(lung_data$Phenotype[temp] == "Other+"), which(is.na(lung_data$Phenotype[temp])))#, which(lung_data$Phenotype[temp] == "CK+"))

if (length(cell_rm) > 0){temp = temp[-cell_rm]
}else {temp = temp}

temp_df = data.frame(x = lung_data$`Cell X Position`[temp], y =  lung_data$`Cell Y Position`[temp], cell_type = as.factor(lung_data$Phenotype[temp]), patient = paste0("Patient ", i))
df <- rbind(df,temp_df)
}

ggplot(df, aes(x = x, y = y, col = cell_type)) + geom_point(size = 0.25) + xlab("X position") + ylab("Y position")   + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + facet_wrap(~ patient)


plt_group = head(worst_surv, 12)
df <- NULL
for(i in plt_group){
   
temp = slide_idx[[i]] 
cell_rm = c(which(lung_data$Phenotype[temp] == "Other+"), which(is.na(lung_data$Phenotype[temp])))#, which(lung_data$Phenotype[temp] == "CK+"))
if (length(cell_rm) > 0){temp = temp[-cell_rm]
}else {temp = temp}

temp_df = data.frame(x = lung_data$`Cell X Position`[temp], y =  lung_data$`Cell Y Position`[temp], cell_type = as.factor(lung_data$Phenotype[temp]), patient = paste0("Patient ", i))
df <- rbind(df,temp_df)
}

ggplot(df, aes(x = x, y = y, col = cell_type)) + geom_point(size = 0.25) + xlab("X position") + ylab("Y position")   + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + facet_wrap(~ patient)
```

```{r, warning=FALSE, echo = FALSE, message=FALSE}
n_break = 20
entropy_all = list()  
for (i in 1:n_patient){
  if(i %% 20 == 0){
    cat(paste0("iteration: ", i, "\n"))
  }
  
  
temp = slide_idx[[i]]

cell_rm = c(which(lung_data$Phenotype[temp] == "Other+"), which(is.na(lung_data$Phenotype[temp])))
#cell_rm = c(which(lung_data$Phenotype[temp] == "Other+"), which(is.na(lung_data$Phenotype[temp])), which(lung_data$Phenotype[temp] == "CK+"))

if (length(cell_rm) > 0){temp = temp[-cell_rm]
}else {temp = temp}


x_pos = lung_data$`Cell X Position`[temp]
y_pos = lung_data$`Cell Y Position`[temp]
patient_i = ppp(x_pos, y_pos, c(min(x_pos), max(x_pos)), c(min(y_pos), max(y_pos)))
patient_i = patient_i %mark% as.factor(lung_data$Phenotype[temp])

# QUANTILES
#dist_i = pairdist(patient_i)
#dist_i = dist_i[lower.tri(dist_i, diag = FALSE)]
#dist_i = quantile(dist_i, probs = prob_seq)

# ADAPTIVE DISTANCE BREAKS
#rmax_i = diameter(as.rectangle(patient_i)) 

dist_i = pairdist(patient_i)
rmax_i= max(dist_i[lower.tri(dist_i, diag = FALSE)])
dist_i = exp(seq(log(0.05*rmax_i), log(rmax_i), length.out = n_break))  #make.even.breaks(rmax_i, bstep = rmax_i/20)$r
dist_i = dist_i[-length(dist_i)]

entropy_all[[i]] = altieri(patient_i, distbreak = dist_i) #dist_range
}

#ttt = pairdist(patient_i)
#min(ttt[lower.tri(ttt, diag = FALSE)])
#max(ttt[lower.tri(ttt, diag = FALSE)])
```
#setwd("~/Documents/Imaging project/Vectra /Spatial entropy output")
#save(entropy_all, file = "lung_maxdist.RData) #based on max pairdist to create sequence
#load("lung_maxdist.RData")
#load("lung_immune_maxdist.RData") # without CK+

```{r, warning=FALSE, echo = FALSE, message=FALSE}
n_patient = length(entropy_all)
dist_all = spi_all = res_all = list()
df_all <- NULL
for(i in 1:n_patient){
   
temp = entropy_all[[i]]
dist_i = unique(c(temp$distance.breaks[,1],temp$distance.breaks[,2]))[-1]
spi_i = temp$SPI.terms
res_i = temp$RES.terms
dist_all[[i]] = dist_i
spi_all[[i]] = spi_i
res_all[[i]] = res_i
temp_df = data.frame(dist_break = dist_i, spi_i = spi_i, res_i  =  res_i, patient = paste0("Patient ", i))
df_all = rbind(df_all, temp_df)
}
```

### figures for manuscript

```{r, warning=FALSE, echo = FALSE, message=FALSE}

# for images 28 vs 137
i = 137 #137
df <- NULL

temp = slide_idx[[i]] 
cell_rm = c(which(lung_data$Phenotype[temp] == "Other+"), which(is.na(lung_data$Phenotype[temp])))#, which(lung_data$Phenotype[temp] == "CK+"))

if (length(cell_rm) > 0){temp = temp[-cell_rm]
}else {temp = temp}

temp_df = data.frame(x = lung_data$`Cell X Position`[temp], y =  lung_data$`Cell Y Position`[temp], cell_type = as.factor(lung_data$Phenotype[temp]))
df <- rbind(df,temp_df)


ggplot(df, aes(x = x, y = y, col = cell_type)) + geom_point(size = 0.6) + xlab("X position") + ylab("Y position")   + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "bottom")
# corresponding spi

temp <- NULL
temp_df <- data.frame(x = dist_all[[i]], y= spi_all[[i]], col=i)
temp <- rbind(temp,temp_df)

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line() +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,1)

# all spi curves

temp <- NULL
for(i in 1:n_patient){ 
  temp_df <- data.frame(x = dist_all[[i]], y= res_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") #+ ylim(0,2.2)

```


```{r, warning=FALSE, echo = FALSE, message=FALSE}
#plt_group = c(125, 91, 47, 112, 139,99)
plt_group = c(73, 88, 100, 96, 103)
df <- NULL
for(i in plt_group){
   
temp = slide_idx[[i]] 
cell_rm = c(which(lung_data$Phenotype[temp] == "Other+"))#, which(is.na(lung_data$Phenotype[temp])), which(lung_data$Phenotype[temp] == "CK+"))

if (length(cell_rm) > 0){temp = temp[-cell_rm]
}else {temp = temp}

temp_df = data.frame(x = lung_data$`Cell X Position`[temp], y =  lung_data$`Cell Y Position`[temp], cell_type = as.factor(lung_data$Phenotype[temp]), patient = paste0("Patient ", i))
df <- rbind(df,temp_df)
}

ggplot(df, aes(x = x, y = y, col = cell_type)) + geom_point(size = 0.25) + xlab("X position") + ylab("Y position")   + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + facet_wrap(~ patient)

temp <- NULL
for(i in plt_group){ 
  temp_df <- data.frame(x = dist_all[[i]], y= res_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,2.5)


```





```{r}
df_all$spi_i[is.infinite(df_all$spi_i)] = 0
df1 = df_all[,-3]
colnames(df1) = c(".index", ".value", ".id")
df1 = df1[,c(".id", ".index", ".value")]

test_mdl = fpca.sc(ydata =df1)
plot_shiny(obj = test_mdl)
```





```{r, warning=FALSE, echo = FALSE, message=FALSE}
#best vs worst survival

temp <- NULL
for(i in best_surv){ 
  temp_df <- data.frame(x = dist_all[[i]], y= spi_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,2.5)

temp <- NULL
for(i in worst_surv){ 
  temp_df <- data.frame(x = dist_all[[i]], y= spi_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,2.5)


temp <- NULL
for(i in best_surv){ 
  temp_df <- data.frame(x = dist_all[[i]], y= res_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,3)

temp <- NULL
for(i in worst_surv){ 
  temp_df <- data.frame(x = dist_all[[i]], y= res_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,3)

```

### FPCA with PACE 
```{r, warning=FALSE, echo = FALSE, message=FALSE}
test_res1 <- FPCA(spi_all, dist_all)#, list(userBwMu = 10, dataType='Sparse', error=FALSE, kernel='epan', verbose=TRUE))
plot(test_res1) 
#test_res2 <- FPCA(res_all, dist_all)#, list(userBwMu = 10, dataType='Sparse', error=FALSE, kernel='epan', verbose=TRUE))
#plot(test_res2) 
# predicted trajectories?
#CreatePathPlot(test_res, subset = c(head(best_surv, 3)), showMean = T, ylim = c(0,2.5))
#CreatePathPlot(test_res, subset = c(head(worst_surv, 3)), showMean = T, ylim = c(0,2.5))

```

```{r}
plot(test_res1$xiEst[,1], test_res1$xiEst[,2])

```


```{r}
CreateCovPlot(test_res1, corr = TRUE)
#CreateCovPlot(test_res2, corr = TRUE)
```

```{r}
# PC score functions
Xi_hat1 = test_res1$xiEst
#Xi_hat2 = test_res2$xiEst

data_combine = data.frame(event = lung_clinical_data$`Censor (0=Alive, 1=Dead)`, stime = lung_clinical_data$`Survival (days)`/365,
                          stage_group = as.factor(lung_clinical_data$`Stage (numeric)`), Age = lung_clinical_data$`Age at Diagnosis`, 
                          spa_e = Xi_hat1)#, res_e = Xi_hat2)  #spa_e: spatial entropy, res_e: residual entropy

# 5 year survival
#event_id = which(data_combine$event == 1)
#data_combine$event[event_id] = ifelse(data_combine$stime[event_id] > 10, 0, 1)

test_fit <- gam(stime ~ Age + stage_group + spa_e.1 + spa_e.2 , weights=event, data = data_combine, family=cox.ph()) #+ PC1.1 + PC1.2

summary(test_fit)
# alternatively
#fullcoxresult<-coxph(formula=Surv(stime,event)~ ., data = data_combine)
fullcox<-coxph(formula=Surv(stime,event)~ Age + stage_group + spa_e.1 + spa_e.2, data = data_combine)
nestcox<-coxph(formula=Surv(stime,event)~ Age + stage_group, data = data_combine)
anova(fullcox, nestcox)
```



```{r}
plot(test_mdl$scores[,1], test_mdl$scores[,2])
```

```{r}
# PC score functions
Xi_hat1 = test_mdl$scores


data_combine = data.frame(event = lung_clinical_data$`Censor (0=Alive, 1=Dead)`, stime = lung_clinical_data$`Survival (days)`/365,
                          stage_group = as.factor(lung_clinical_data$`Stage (numeric)`), Age = lung_clinical_data$`Age at Diagnosis`, 
                          spa_e = Xi_hat1)
# 5 year survival
#event_id = which(data_combine$event == 1)
#data_combine$event[event_id] = ifelse(data_combine$stime[event_id] > 10, 0, 1)

test_fit <- gam(stime ~ Age + stage_group + spa_e.1 + spa_e.2 + spa_e.3 , weights=event, data = data_combine, family=cox.ph()) 

summary(test_fit)
# alternatively
#fullcoxresult<-coxph(formula=Surv(stime,event)~ ., data = data_combine)
# alternatively
fullcox<-coxph(formula=Surv(stime,event)~ Age + stage_group + spa_e.1 + spa_e.2 + spa_e.3, data = data_combine)
nestcox<-coxph(formula=Surv(stime,event)~ Age + stage_group, data = data_combine)
anova(fullcox, nestcox)
```

###################################
##TNBC
##################################

```{r, warning=FALSE}
# TNBC data
setwd("/Users/vut3/Documents/Imaging project/Vectra /TNBC_shareCellData")
#save.image(file = "tnbc_workspace.RData")
load("tnbc_workspace.RData")
```


```{r}
tnbc = tnbc %>% mutate("Phenotype" = case_when(`Group` == 6 | `Group` == 5 ~ "Tumor", `Group` == 1  ~ "Unidentified", `Group` == 2  ~ "Immune", `Group` == 3  ~ "Endothelial", `Group` == 4  ~ "Mesenchymal-like"))

```

```{r}
samp_rm = unique(c(which(tnbc_class$patient_class == 2), 22, 30, 38))
n_surv = 15
event_id = which(tnbc_clinical$Censored == 0)
censor_id = which(tnbc_clinical$Censored == 1)
#best_surv = event_id[order(lung_clinical_data$`Survival (days)`[event_id], decreasing = T)[1:n_surv]]
best_surv = censor_id[order(tnbc_clinical$`Survival_days_capped*`[censor_id], decreasing = T)[1:n_surv]]
worst_surv = event_id[order(tnbc_clinical$`Survival_days_capped*`[event_id])[1:n_surv]]
```


```{r}
entropy_tnbc = list()
for (i in 1:N){
  if(i %% 20 == 0){
    cat(paste0("iteration: ", i, "\n"))
  }
  
  if(i == 30){next}
  else{
  
    
  temp = all_idx[[i]]
  
  #remove unidentified cells
  #cell_rm = which(tnbc$Group[temp] == 1)
   cell_rm = which(tnbc$Phenotype[temp] == "Unidentified")
  if (length(cell_rm) > 0){
    temp = temp[-cell_rm]
    x_pos = center_x[[i]][-cell_rm]
    y_pos = center_y[[i]][-cell_rm]
  }else {temp = temp
  x_pos = center_x[[i]]
  y_pos = center_y[[i]]
  }
  patient_i = ppp(x_pos, y_pos, c(min(x_pos), max(x_pos)), c(min(y_pos), max(y_pos))) 
  patient_i = patient_i %mark% as.factor(tnbc$Phenotype[temp]) #as.factor(tnbc$Group[temp])
  
  dist_i = pairdist(patient_i)
  rmax_i= max(dist_i[lower.tri(dist_i, diag = FALSE)])
  dist_i = exp(seq(log(0.05*rmax_i), log(rmax_i), length.out = 20))  #make.even.breaks(rmax_i, bstep = rmax_i/20)$r
  dist_i = dist_i[-length(dist_i)]
 
   entropy_tnbc[[i]] = altieri(patient_i, distbreak = dist_i) 
  }
}
#ttt = pairdist(patient_i)
#min(ttt[lower.tri(ttt, diag = FALSE)])
#max(ttt[lower.tri(ttt, diag = FALSE)])
```
#setwd("~/Documents/Imaging project/Vectra /Spatial entropy output")
#load("tnbc_maxdist.RData")
```{r}
dist_all = spi_all = res_all = list()
df <- NULL
for(i in 1:n_patient){
   
  
if(i == 30){next}
  else{  
temp = entropy_tnbc[[i]]
dist_i = unique(c(temp$distance.breaks[,1],temp$distance.breaks[,2]))[-1]
spi_i = temp$SPI.terms
res_i = temp$RES.terms
dist_all[[i]] = dist_i * 0.39
spi_all[[i]] = spi_i
res_all[[i]] = res_i
temp_df = data.frame(dist_break = dist_i * 0.39, spi_i = spi_i, res_i  =  res_i, patient = paste0("Patient ", i))
df = rbind(df, temp_df)
  }
}
```

```{r}
# figs for manuscript
plt_group = c(1:4)
df1 <- NULL
for(i in plt_group){
  temp = all_idx[[i]]
#id = which(new_df$SampleID == i)

 cell_rm = which(tnbc$Phenotype[temp] == "Unidentified")
  if (length(cell_rm) > 0){
    temp = temp[-cell_rm]
    x_pos = center_x[[i]][-cell_rm]
    y_pos = center_y[[i]][-cell_rm]
  }else {temp = temp
  x_pos = center_x[[i]]
  y_pos = center_y[[i]]
  }
 
 
temp = data.frame(x = x_pos, y =  y_pos, cell_type = as.factor(tnbc$Phenotype[temp]), patient = i)
df1 = rbind(df1, temp)
}


ggplot(df1, aes(x = x, y = y, col = cell_type)) + geom_point(size = 0.1) + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + facet_wrap(~ patient, scales = "free") + guides(x = "none", y = "none") #xlab("X position") + ylab("Y position") 


ggplot(df, aes(x = dist_break, y = spi_i, group = patient)) + geom_line(alpha = 0.4) + xlab("r (microns)") + ylab("SPI")   + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none")


```


```{r, warning=FALSE, echo = FALSE, message=FALSE}

ggplot(df, aes(x = dist_break, y = spi_i, col = patient)) + geom_line() + xlab("r (microns)") + ylab("Spatial Entropy")   + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none")

ggplot(df, aes(x = dist_break, y = res_i, col = patient)) + geom_line() + xlab("r (microns)") + ylab("Residual Entropy")   + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none")

```

#best_surv vs worst_surv
```{r}
temp <- NULL
for(i in best_surv){ 
  temp_df <- data.frame(x = dist_all[[i]], y= spi_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,1)


temp <- NULL
for(i in c(23,16,7,21,2,8,13)){ 
  temp_df <- data.frame(x = dist_all[[i]], y= spi_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,1)
```



### FPCA with PACE 
```{r, warning=FALSE, echo = FALSE, message=FALSE}
test_res1 <- FPCA(spi_all[-samp_rm], dist_all[-samp_rm])#, list(userBwMu = 10, dataType='Sparse', error=FALSE, kernel='epan', verbose=TRUE))
plot(test_res1) 
#test_res2 <- FPCA(res_all[-samp_rm], dist_all[-samp_rm])#, list(userBwMu = 10, dataType='Sparse', error=FALSE, kernel='epan', verbose=TRUE))
#plot(test_res2) 
# predicted trajectories?
#CreatePathPlot(test_res1, subset = c(head(best_surv, 3)), showMean = T, ylim = c(0,2.5))
#CreatePathPlot(test_res1, subset = c(head(worst_surv, 3)), showMean = T, ylim = c(0,2.5))
```

```{r}
# plot mean and eigenfunctions
mu_plot = data.frame(x = test_res1$workGrid, y = test_res1$mu)
ggplot(mu_plot, aes(x = x, y = y)) + geom_line() +  xlab("Distance range") + ylab("Mean Function") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") 

phi_plot = NULL
for (i in 1:2){
temp = data.frame(x = test_res1$workGrid, y = test_res1$phi[,i], idx = i)
phi_plot = rbind(phi_plot, temp)
}

ggplot(phi_plot, aes(x = x, y = y, group = idx, col = factor(idx))) + geom_line() +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "bottom") + labs(col = "Eigenfunction")

```


```{r}
CreateCovPlot(test_res1, corr = TRUE)
#CreateCovPlot(test_res2, corr = TRUE)
```

```{r}
# PC score functions
Xi_hat1 = test_res1$xiEst
#Xi_hat2 = test_res2$xiEst


test_df = tnbc_clinical
clin_rm = unlist(sapply(samp_rm, function(x) which(tnbc_clinical$InternalId == x))) 

test_df = test_df[-clin_rm,]
test_df$patient_class = tnbc_class$patient_class[-samp_rm]

data_combine = data.frame(event = ifelse(test_df$Censored == 1, 0, 1), stime = test_df$`Survival_days_capped*`/365, pclass = factor(test_df$patient_class),  Age = test_df$AGE_AT_DX, spa_e = Xi_hat1)#, res_e = Xi_hat2)  #spa_e: spatial entropy, res_e: residual entropy


test_fit <- gam(stime ~   Age + spa_e.1 , weights=event, data = data_combine, family=cox.ph()) 

summary(test_fit)

# alternatively
#fullcoxresult<-coxph(formula=Surv(stime,event)~ ., data = data_combine)
#fullcoxresult<-coxph(formula=Surv(stime,event)~ pclass, data = data_combine)

```


```{r}
# alternatively
fullcox<-coxph(formula=Surv(stime,event)~ Age + spa_e.1 + spa_e.2 , data = data_combine) 
nestcox<-coxph(formula=Surv(stime,event)~ Age  , data = data_combine)
anova(fullcox, nestcox)
```
#=====================#
### ovarian data
#=====================#
```{r, warning=FALSE, message=FALSE}

setwd("~/Documents/Imaging project/Vectra /Ovarian_survival_output")
load("Ovarian_workspace.RData")
```


```{r, warning=FALSE, message=FALSE}
ovarian_data = ovarian_data %>% mutate(`Phenotype CD3+CD8+`=case_when(
  `Phenotype CD3` =="CD3+" & `Phenotype CD8`=="CD8+" ~ "CD3+CD8+",
  TRUE~ ""
))


ovarian_data = ovarian_data %>% mutate(`Phenotype CD3+CD8-`=case_when(
  `Phenotype CD3` =="CD3+" & `Phenotype CD8`=="CD8-" ~ "CD3+CD8-",
  TRUE~ ""
))
 

ovarian_data = ovarian_data %>% mutate(`Phenotype all` = case_when(
  `Phenotype CD3+CD8+` == "CD3+CD8+"  ~ "CD8+",
  `Phenotype CD3+CD8-` == "CD3+CD8-" ~ "CD4+",
  `Phenotype CD19` == "CD19+" ~ "CD19+",
  `Phenotype CD68` == "CD68+" ~ "CD68+",
  `Phenotype CK` == "CK+" ~ "CK+",
  TRUE~""
))

#ovarian_data$`Phenotype all` = NULL

```

#best vs worst survival 
```{r}
n_surv = 15
event_id = which(clinical_data_merge2$`Death (1=yes, 0=no)` == 1)
censor_id = which(clinical_data_merge2$`Death (1=yes, 0=no)` == 0)
#best_surv = event_id[order(lung_clinical_data$`Survival (days)`[event_id], decreasing = T)[1:n_surv]]
best_surv = censor_id[order(clinical_data_merge2$stime[censor_id], decreasing = T)[1:n_surv]]
worst_surv = event_id[order(clinical_data_merge2$stime[event_id])[1:n_surv]]

#plt_group =  c(1,73,  69,100, 120, 118)
patient_idx1 = patient_idx[image_idx]

df <- NULL
for(i in sort(head(best_surv))){
 temp = patient_idx1[[i]]
 cell_rm = c(which(ovarian_data$`Phenotype all`[temp] == ""), which(ovarian_data$`Phenotype all`[temp] == "CK+"))
     #which(ovarian_data$`Phenotype all`[temp] == "")#
 if (length(cell_rm) > 0){temp = temp[-cell_rm]
}else {temp = temp}

temp_df = data.frame(x = ovarian_data$`Cell X Position`[temp], y =  ovarian_data$`Cell Y Position`[temp], cell_type = as.factor(ovarian_data$`Phenotype all`[temp]), patient = paste0("p", i))
df <- rbind(df,temp_df)

}

df$patient = factor(df$patient, levels = unique(df$patient))
ggplot(df, aes(x = x, y = y, col = cell_type)) + geom_point(size = 0.1)  + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + facet_wrap(~ patient, scales = "free")+ guides(x = "none", y = "none") #xlab("X position") + ylab("Y position") 


df <- NULL
for(i in sort(head(worst_surv))){
   
 temp = patient_idx1[[i]]
 cell_rm = c(which(ovarian_data$`Phenotype all`[temp] == ""), which(ovarian_data$`Phenotype all`[temp] == "CK+"))
     #which(ovarian_data$`Phenotype all`[temp] == "")#
 if (length(cell_rm) > 0){temp = temp[-cell_rm]
}else {temp = temp}

temp_df = data.frame(x = ovarian_data$`Cell X Position`[temp], y =  ovarian_data$`Cell Y Position`[temp], cell_type = as.factor(ovarian_data$`Phenotype all`[temp]), patient = paste0("p", i))
df <- rbind(df,temp_df)

}
df$patient = factor(df$patient, levels = unique(df$patient))
ggplot(df, aes(x = x, y = y, col = cell_type)) + geom_point(size = 0.1) + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + facet_wrap(~ patient, scales = "free") #+ guides(x = "none", y = "none") #xlab("X position") + ylab("Y position") 

# for manuscript

df <- NULL
for(i in c(9, 25, 90, 102)){
   
 temp = patient_idx1[[i]]
 cell_rm = c(which(ovarian_data$`Phenotype all`[temp] == ""), which(ovarian_data$`Phenotype all`[temp] == "CK+"))
     #which(ovarian_data$`Phenotype all`[temp] == "")#
 if (length(cell_rm) > 0){temp = temp[-cell_rm]
}else {temp = temp}

temp_df = data.frame(x = ovarian_data$`Cell X Position`[temp], y =  ovarian_data$`Cell Y Position`[temp], cell_type = as.factor(ovarian_data$`Phenotype all`[temp]), patient = paste0("p", i))
df <- rbind(df,temp_df)

}
df$patient = factor(df$patient, levels = unique(df$patient))
ggplot(df, aes(x = x, y = y, col = cell_type)) + geom_point(size = 0.1) + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + facet_wrap(~ patient, scales = "free") + guides(x = "none", y = "none") #xlab("X position") + ylab("Y position") 

```


```{r, warning=FALSE, echo = FALSE, message=FALSE}
# spatial entropy and residual entropy
n_break = 20
ovarian_entropy = list()
#dist_range = c(25, 50, 80, 100, 120, 150, 200, 300, 400, 500, 600, 700, 1000) 

for (i in 1:n_patient){
  if(i %% 20 == 0){
    cat(paste0("iteration: ", i, "\n"))
  }
  
temp = patient_idx[[i]]

cell_rm = c(which(ovarian_data$`Phenotype all`[temp] == ""), which(ovarian_data$`Phenotype all`[temp] == "CK+"))
if (length(cell_rm) > 0){temp = temp[-cell_rm]
}else {temp = temp}

x_pos = ovarian_data$`Cell X Position`[temp]
y_pos = ovarian_data$`Cell Y Position`[temp]
patient_i = ppp(x_pos, y_pos, c(min(x_pos), max(x_pos)), c(min(y_pos), max(y_pos)))
patient_i = patient_i %mark% as.factor(ovarian_data$`Phenotype all`[temp])


dist_i = pairdist(patient_i)
rmax_i= max(dist_i[lower.tri(dist_i, diag = FALSE)])
dist_i = exp(seq(log(0.05*rmax_i), log(rmax_i), length.out = 20))  #make.even.breaks(rmax_i, bstep = rmax_i/20)$r
dist_i = dist_i[-length(dist_i)]
 

ovarian_entropy[[i]] = altieri(patient_i, distbreak = dist_i) 
}
```

#setwd("~/Documents/Imaging project/Vectra /Spatial entropy output")
#load("ovarian_maxdist.RData")
```{r}

ovarian_entropy = ovarian_entropy[image_idx]
n_break = length(ovarian_entropy[[1]]$class.pairs) #20
dist_all = spi_all = res_all = list()
df <- NULL
for(i in 1:length(ovarian_entropy)){
   
temp = ovarian_entropy[[i]]
dist_i = unique(c(temp$distance.breaks[,1],temp$distance.breaks[,2]))[-1]
spi_i = temp$SPI.terms
res_i = temp$RES.terms
dist_all[[i]] = dist_i 
spi_all[[i]] = spi_i
res_all[[i]] = res_i
temp_df = data.frame(dist_break = dist_i, spi_i = spi_i, res_i  =  res_i, patient = paste0("Patient ", i))
df = rbind(df, temp_df)
}
```


```{r}
#unlist(lapply(spi_all, function(x) sum(is.finite(x))))
# there're some inf values in spi_all: #44, 69
#entropy calculation
e_spatial_calc = function(p1, p2){
  # p1, p2: vector of probabilities numer vs. denom
  idx1 = which(p1 == 0)
  idx2 = which(p2 == 0)
  id_rm = union(idx1, idx2)
  
  res = sum(p1[-id_rm]*log(p1[-id_rm]/p2[-id_rm]))
  
  return(res)
}


check_group = which(unlist(lapply(spi_all, function(x) sum(is.finite(x)))) != n_break)
for (i in check_group){
temp = lapply(ovarian_entropy[[i]]$cond.Z.distribution, function(x) e_spatial_calc(x$rel.freq, ovarian_entropy[[i]]$ShannonZ$`marginal distribution`$rel.freq))
spi_all[[i]] = as.numeric(unlist(temp))
}
```

```{r}
df <- NULL
for(i in 1:length(ovarian_entropy)){
dist_i = dist_all[[i]] 
spi_i = spi_all[[i]] 
res_i = res_all[[i]]
temp_df = data.frame(dist_break = dist_i, spi_i = spi_i, res_i  =  res_i, patient = paste0("Patient ", i))
df = rbind(df, temp_df)
}
```

```{r, warning=FALSE, echo = FALSE, message=FALSE}

ggplot(df, aes(x = dist_break, y = spi_i, group = patient)) + geom_line(alpha = 0.4) + xlab("r (microns)") + ylab("Spatial Entropy")   + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none")


ggplot(df, aes(x = dist_break, y = res_i, col = patient)) + geom_line() + xlab("r (microns)") + ylab("Residual Entropy")   + theme(plot.title = element_text(hjust = 0.5))+ theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none")

```


```{r}
temp <- NULL
for(i in best_surv){ 
  temp_df <- data.frame(x = dist_all[[i]], y = spi_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,1)


temp <- NULL
for(i in worst_surv){ 
  temp_df <- data.frame(x = dist_all[[i]], y = spi_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") + ylim(0,1)

temp <- NULL
for(i in best_surv){ 
  temp_df <- data.frame(x = dist_all[[i]], y = res_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") 


temp <- NULL
for(i in worst_surv){ 
  temp_df <- data.frame(x = dist_all[[i]], y = res_all[[i]], col=i)
  temp <- rbind(temp,temp_df)} 

ggplot(temp,aes(x = x, y = y, group=col)) + geom_line(stat = "identity", alpha = 0.4) +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") 

```


```{r}

df1 = df[,-3]
colnames(df1) = c(".index", ".value", ".id")
df1 = df1[,c(".id", ".index", ".value")]

test_mdl = fpca.sc(ydata = df1)
#plot_shiny(obj = test_mdl)
```


```{r}
plot(test_mdl$scores[,1:2])
```

```{r, warning=FALSE, echo = FALSE, message=FALSE}
# spatial entropy has inf values due to 0 prob in marginal Z distribution
test_res1 <- FPCA(spi_all, dist_all)#, list(userBwMu = 10, dataType='Sparse', error=FALSE, kernel='epan', verbose=TRUE))
plot(test_res1) 
test_res2 <- FPCA(res_all, dist_all)#, list(userBwMu = 10, dataType='Sparse', error=FALSE, kernel='epan', verbose=TRUE))
plot(test_res2) 
test_res1$cumFVE
# predicted trajectories?
#CreatePathPlot(test_res1, subset = c(head(best_surv, 3)), showMean = T, ylim = c(0,2.5))
#CreatePathPlot(test_res1, subset = c(head(worst_surv, 3)), showMean = T, ylim = c(0,2.5))

#CreateCovPlot(test_res1, corr = TRUE)
```

```{r}
# plot mean and eigenfunctions
mu_plot = data.frame(x = test_res1$workGrid, y = test_res1$mu)
ggplot(mu_plot, aes(x = x, y = y)) + geom_line() +  xlab("Distance range") + ylab("Mean Function") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "none") 

phi_plot = NULL
for (i in 1:3){
temp = data.frame(x = test_res1$workGrid, y = test_res1$phi[,i], idx = i)
phi_plot = rbind(phi_plot, temp)
}

ggplot(phi_plot, aes(x = x, y = y, group = idx, col = factor(idx))) + geom_line() +  xlab("Distance range") + ylab("Entropy") + theme_bw() + theme(panel.border = element_blank(), panel.grid.major = element_blank(), panel.grid.minor = element_blank(), axis.line = element_line(colour = "black")) + theme(legend.position = "bottom") + labs(col = "Eigenfunction")

```



```{r}
# PC score functions
Xi_hat1 = test_res1$xiEst #test_mdl$scores[,1:3]
Xi_hat2 = test_res2$xiEst
#plot(Xi_hat1[,1], Xi_hat1[,2])
data_combine = data.frame(event = clinical_data_merge2$`Death (1=yes, 0=no)`, stime = clinical_data_merge2$stime, 
                          Age = clinical_data_merge2$`Age at Diagnosis`,  spi_e = Xi_hat1) #, res_e = Xi_hat2)  #spa_e: spatial entropy, res_e: residual entropy

test_fit <- gam(stime ~   Age + spi_e.1 + spi_e.2 + spi_e.3 , weights=event, data = data_combine, family=cox.ph()) #
test_fit2  = gam(stime ~   Age , weights=event, data = data_combine, family=cox.ph()) 
summary(test_fit)
# alternatively
fullcox<-coxph(formula=Surv(stime,event)~ Age + spi_e.1 + spi_e.2 + spi_e.3 , data = data_combine) #
nestcox<-coxph(formula=Surv(stime,event)~ Age  , data = data_combine)
anova(fullcox, nestcox)
```


